---
title: "Unraveling Credit Risk: Inference and Prediction Models for S&P 500 Bonds"
author: "Kelly Tong, Rakeen Rouf, Lisa Wang, Javier Cervantes"
date: 11/28/2023
abstract: This analysis delves into crucial dimensions of credit risk using a dataset gathered on September 22, 2023. Focused on bonds issued exclusively by S&P 500-listed companies, the dataset encompasses vital information, including bond details, company fundamentals, credit ratings, and social sentiment indicators. The research tackles two pivotal questions -- the determinants of a bond's credit rating and the predictability of credit spreads. Leveraging ordinal regression and linear regression models, the study reveals insights into credit risk assessment and introduces a model with 67% RMSE in predicting credit spreads. This comprehensive exploration provides valuable implications for investors and financial analysts navigating the intricacies of credit risk within the dynamic financial landscape.
format:
  pdf:  
    mainfont: Helvetica Neue  
    fontsize: 10pt
    papersize: A4
    margin-top: 20mm
    margin-bottom: 20mm  
    margin-left: 20mm
    margin-right: 20mm
    number-sections: true
    number-offset: 3
execute:
  echo: false
  warning: false
editor: visual
---

# Introduction

In the complex landscape of finance, the management and assessment of credit risk play pivotal roles in shaping investment strategies and influencing market dynamics. The finance problem at the heart of this research revolves around the need for a deeper understanding of credit risk within the context of bonds issued by companies listed in the S&P 500. As investors seek avenues to optimize their portfolios and financial institutions strive for effective risk management, the questions of what factors contribute to a bond's credit rating and how credit spreads can be predicted become paramount. In an ever-evolving financial environment, where market conditions and sentiment can swiftly impact investment outcomes, addressing these questions is not merely an academic pursuit but a practical necessity. The outcomes of this analysis hold the potential to refine credit risk assessment methodologies, offering tangible benefits for investors, financial analysts, and the broader financial ecosystem.

## Data Overview

The dataset we used to analyze the research problem is a subset of the holdings within an ETF, exclusively representing companies listed in the S&P 500. It comprises **2,341** rows, with each row corresponding to a specific bond issued by an S&P 500 company. Across the dataset, there are **34** variables, which can be grouped into four distinct categories:

1.  **Bond information from iShare:** Information related to the bonds, including the issuer's name, industry sector, price, duration, yield to maturity, issuer's stock ticker, and market capitalization. Sourced from the USIG Ishares Credit Bond ETF[^1]

2.  **Company fundamentals from Yahoo Finance:** Company fundamentals, including various financial ratios (e.g., revenue, debt). Sourced from Yahoo Finance[^2] using the yfinance package[^3].

3.  **Credit ratings from Bloomberg:** Credit ratings from Fitch, Moody's, and S&P, and a composite credit rating. Sourced from the Bloomberg Terminal[^4]

4.  **Social sentiment indicators from Finhubb API:** Social sentiment indicators including the number of positive and negative mentions on Reddit last year. Sourced from the Finnhub API[^5]

[^1]: iShares by BlackRock. (2023). Retrieved from https://www.ishares.com/us

[^2]: iShares by BlackRock. (2023). Retrieved from https://www.ishares.com/us

[^3]: Yahoo! Finance. (2023). Company profiles, income statements, balance sheets, and event/announcement details for various companies. Retrieved from http://finance.yahoo.com/

[^4]: Bloomberg L.P. (2023). Retrieved from Bloomberg terminal https://www.bloomberg.com/professional/solution/bloomberg-terminal/

[^5]: Finnhub API. (Version 1.0.0). (2023). Retrieved from https://finnhub.io/docs/api/introduction

\newpage

### Exploratory Data Analysis (EDA) and Data Manipulation

After conducting an exploratory data analysis (EDA) on our original dataset (please refer to our EDA for reference), we have identified several issues that require attention:

### Missing Values

Given the different accounting practices of the different kinds of businesses in each of the Sectors in our universe, there are bound to be many (often very concentrated) NAs. Our approach consists in exploring the nature of those NAs to determine if those values should be removed, replaced or handled in any other way.

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(caret)
library(ggplot2)
library(corrplot)
library(magrittr)
library(knitr)
library(gridExtra)
library(gtsummary)
suppressMessages(library(tidyverse))
```

```{r}
bonds <- read.csv("credit_risk_data.csv")

```

### Colinearity

Many of the fundamental variables included in our model capture similar relationships and could therefore lead to increased variance in our model. To address this, we will perform a Variance Inflation Factor (VIF) analysis on the model. This analysis will assist in evaluating whether certain variables must be removed.

### Outliers

The dataset may contain influential points that can significantly impact our estimators. We plan to perform Cook's Distance analysis to determine if any such points exist.

### Sector & Credit Ratings

```{r}
# collapsing credit ratings
bonds$BB_COMPOSITE <- ifelse(bonds$BB_COMPOSITE %in% c("AAA", "AA+", "AA", "AA-"), ">A+", bonds$BB_COMPOSITE)
bonds$BB_COMPOSITE <- ifelse(bonds$BB_COMPOSITE %in% c('BB+', 'BBB-'), "<BBB", bonds$BB_COMPOSITE)
bonds$BB_COMPOSITE <- factor(bonds$BB_COMPOSITE, levels = c('>A+', 'A+', 'A', 'A-', 'BBB+', 'BBB', '<BBB'))
```

```{r}
# Collapsing categories in Sector.
bonds$Sector <- ifelse(bonds$Sector == "Brokerage/Asset Managers/Exchanges", "Finance", bonds$Sector)
bonds$Sector <- ifelse(bonds$Sector == "Electric", "Energy", bonds$Sector)
bonds$Sector <- ifelse(bonds$Sector == "Natural Gas", "Energy", bonds$Sector)
bonds$Sector <- ifelse(bonds$Sector == "Reits", "Finance", bonds$Sector)
bonds$Sector <- ifelse(bonds$Sector == "Transportation", "Communications", bonds$Sector)
```

There is a multitude of categories in Credit Ratings and Sector, with some having minimal data. This situation could lead to high standard errors and adversely affect the quality of our model. Our strategy is to combine categories with limited data into broader categories, addressing the potential issues related to low data volume.

By addressing these concerns through careful data manipulation, we aim to enhance the quality and reliability of our dataset for subsequent analyses.

## Models

\*\* include a description of the models that will be used for each research question

\*\* include material from the statistical analysis plan

For each of the two research question, we are going to fit three models and compare the differences between them to better understand the problem. The first model will be a full model with all the variable we selected. The second model will be limited to fewer variables. The third model will also include interaction terms.

### Ordinal Credit Score Model Performance

In our pursuit of building an ordinal model to infer credit scores, we have fitted three distinct models. To gauge their performance, we employ the Akaike Information Criterion (AIC) score for comparison. Additionally, we assess the model on the training dataset by generating a confusion matrix. This analysis provides insights into how well our model aligns with the actual data. Crucially, we scrutinize the coefficients to understand the relationships between predictors and outcomes, seeking valuable insights from the model.

### Linear Credit Spread Model Performance

Given the constraints of limited data, we opt for a 10-fold cross-validation approach to fit the three proposed linear models for predicting credit spreads. This involves partitioning the dataset into 10 folds and training the models iteratively. To ensure replicability, we introduce a random seed for randomized fold selection. Subsequently, we employ the Root Mean Square Error (RMSE) metric to evaluate model performance. The model with the lowest RMSE is chosen, indicating the most favorable predictive accuracy. This rigorous process allows us to select a robust linear model for credit spread prediction.

## Results

### Handling NAs

**Coverage ratios:** EDA suggests that the coverage ratios show no meaningful relationship with our outcome variables in addition to containing a large amount of NAs. The reason for this could be the different accounting practices and the different nature of the businesses in our universe. Of particular relevance is the different treatment of Interest Expense as an accounting entry which is a main component in the calculation of these ratios. Since these predictors will surely result in added variance and noise, they have been removed from our model.

```{r, include=FALSE}
# int_coverage 
int_coverage_nas <- bonds[is.na(bonds$int_coverage),] %>% count(Sector)

int_coverage_nas <- merge(int_coverage_nas, bonds %>% count(Sector), by = "Sector")

colnames(int_coverage_nas) <- c('Sector', 'NA_count', 'total_count') 
print(int_coverage_nas)
```

**Liquidity ratios:** Once again, different accounting practices lead to results that suggest that no meaningful relationship exists between these ratios and our outcome variables. This time, the difference lies in the way different sectors consider the Current Liabilities and Current Assets accounting entries. Of particular relevance are the Banking and Insurance sectors whose Current Assets aren't easily convertible into cash which makes liquidity harder to evaluate.

```{r, include=FALSE}
# cash_ratio
cash_ratio_nas <- bonds[is.na(bonds$cash_ratio),] %>% count(Sector) 
cash_ratio_nas <- merge(cash_ratio_nas, bonds %>% count(Sector), by = "Sector")
colnames(cash_ratio_nas) <- c('Sector', 'NA_count', 'total_count') 
print(cash_ratio_nas)
```

## Diagnostic Plots

### Continuous model

-   Full model:

```{r}
credit_spread_full <- lm(credit_spread ~ Sector + 
                          Duration + 
                          marketCapitalization + 
                          BB_COMPOSITE + 
                          debt_to_assets + 
                          debt_to_equity + 
                          int_coverage + 
                          cash_coverage + 
                          current_ratio + 
                          cash_ratio + 
                          roa + 
                          ebitda_margin + 
                          debt_service_coverage  + 
                          score + 
                          score * Sector + 
                          debt_to_assets * Sector, 
                        data=bonds) 
summary(credit_spread_full)
```

-   Second Model, Variable Selection: RFE

```{r}
credit_spread_red <- lm(credit_spread ~ Sector + 
                          Duration + 
                          marketCapitalization + 
                          BB_COMPOSITE + 
                          debt_to_assets + 
                          roa + 
                          ebitda_margin +  
                          operating_profit_margin + 
                          score, 
                        data=bonds)
summary(credit_spread_red)
```

-   Third Model:

```{r}
credit_spread_com <- lm(credit_spread ~ Sector + 
                          Duration + 
                          marketCapitalization + 
                          BB_COMPOSITE + 
                          roa + 
                          ebitda_margin + 
                          debt_to_assets + 
                          operating_profit_margin + 
                          score + 
                          score * Sector + 
                          debt_to_assets * Sector, 
                        data=bonds) 
summary(credit_spread_com)
```

From the study of the different diagnostic plots we can evaluate if our model's assumptions are being violated.

-   **Linearity**: the residuals vs fitted plot shows no signs of non-linearity

    ```{r, warning=FALSE}
    plot(credit_spread_com, which = 1)
    ```

```{=html}
<!-- -->
```
-   **Normality of errors**: by studying the Q-Q plot, we can observe that the errors don't quite follow a Normal Distribution

    ```{r, warning = FALSE, message = FALSE}
    plot(credit_spread_com, which = 2)
    ```

-   **Influential points**: by studying Cook's distance we found no influential points. Not a single observation went above 0.03

```{r}
# residual plots & cook's distance
plot(credit_spread_com, which = 4)
```

## Variable Selection

-   Colinearity Analysis : placeholder for vif values' table

## Model Fit

### First model

```{r}
library(MASS)
library(tidyverse)
```

```{r}
ord_mod <- polr(BB_COMPOSITE ~ score + Duration + credit_spread + debt_to_assets, data=bonds, Hess=TRUE)
summary(ord_mod)
```

```{r}
pvals <- pnorm(-abs(summary(ord_mod)$coef[,"t value"]))*2
ctable <- cbind(summary(ord_mod)$coef,pvals)

ctable
```

```{r}
exp_coefs <- exp(cbind(OR=coef(ord_mod),confint(ord_mod)))
print(exp_coefs)
```

# Conclusion

## Findings

## Limitation

1.  **Dependence Among Rows:** One notable limitation of our analysis arises from the potential lack of independence among rows. Since multiple rows correspond to bonds issued by the same company, there may be interdependence among observations. A hierarchical modeling approach, incorporating random effects for companies, could address this issue by capturing the inherent correlation within the data
2.  **Generalization to Broader Market Dynamics:** Our analysis focuses on bonds issued by companies listed in the S&P 500, limiting the generalizability of our findings to a broader market context. The dynamics of credit risk assessment may differ for companies outside this scope, and caution should be exercised when extending our conclusions to a more diverse range of entities.
3.  **Non-Time Series Analysis:** It's essential to highlight that our study does not employ a time series analysis. This limitation restricts our ability to capture temporal dynamics and trends in credit risk over time. Future research endeavors could delve into time-series methodologies, providing a more comprehensive understanding of how credit risk evolves and responds to varying economic conditions.

## Future Work

1.  **Machine Learning Model Refinement:** Delving into advanced machine learning algorithms, exploring ensemble methods, or tailoring deep learning architectures specifically for credit risk assessment holds the potential to uncover novel avenues for improving predictive accuracy and unlocking additional dimensions of insight.
2.  **Incorporate External Economic Indicators:** Integrate external economic indicators, such as interest rates, inflation, and GDP growth, into the analysis. Explore their impact on credit risk and assess how macroeconomic conditions influence the creditworthiness of companies and the corresponding bond markets.
3.  **Stakeholder Engagement and User Feedback:** Engage with stakeholders, including investors, financial institutions, and industry experts, to gather feedback on the analysis outcomes. Understand user needs and perspectives to tailor future research efforts to address practical challenges faced by the financial community.
