---
title: "ordinal_regression"
author: "Raks"
format: pdf
editor: visual
---

## Overview

The goal of this analysis is to find out which factors contribute to the credit rating of a company. This analysis is based on a subset of the data described in the data overview section. The number of observations are relatively evenly separated in 7 ordered categories of credit ratings. Therefore, a Ordinal Logistic Regression model is used to answer the research question.

## A priori Variable Selection

`Sentiment Score`: Public's sentiment, as captured by a formulaic score produced using the number of positive and negative mentions on Reddit in the last one year. It is reasonable to assume that the more positive the public's sentiment is of a company, the more likely they are to have a better credit rating.

`Duration`: Duration in investments quantifies the degree to which a bond's or portfolio's price is affected by shifts in interest rates. It is reasonable to assume that the higher the duration of the company, the less its credit rating is likely to be.

`Debt to assest ratio`: The debt-to-assets ratio is a financial metric that compares a company's total debt to its total assets. It is a measure of financial leverage and provides insight into the proportion of a company's assets that are financed by debt. It is reasonable to assume that the lower the debt to asset ratio can cause higher credit rating.

`Debt to equity ratio`: Debt-to-equity ratio is a financial metric that compares a company's total debt to its total equity. It is calculated by dividing the company's total debt by its shareholders' equity. The debt-to-equity ratio is a commonly used measure of a company's financial leverage and indicates the proportion of a company's funding that comes from debt compared to equity. It is reasonable to assume that a company with a lower Debt to equity ratio, will have higher a credit rating.

`Credit spread`: Credit spread represents the additional yield or interest rate demanded by investors to hold a bond with a higher credit risk compared to a risk-free asset, such as a government bond. It is reasonable to assume that a company with lower credit spread will have a higher credit rating.

`Operating profit margin`: Operating profit margin is a key financial ratio that represents the percentage of revenue that remains after operating expenses are deducted. It is reasonable to assume that a company withg a higher operating profit margin will have a higher credit rating.

`EBIDTA margin`: EBITDA margin is a measure of a company's operating profitability before accounting for interest, taxes, and non-cash items. It is reasonable to assume that a company with a higher EBIDTA margin will have a better credit rating.

`Return on Assets (ROA)`: Return on Assets (ROA) is a financial metric that measures a company's profitability by expressing its net income as a percentage of its total assets. It is reasonable to assume that a higher ROA, will cause a better credit rating.

`Interest Coverage raito`: The interest coverage ratio is a financial metric that assesses a company's ability to meet its interest payments on outstanding debt. It is reasonable to assume that a company with a higher interest coverage ratio will cause a better credit rating.

## Data Cleaning

```{r}
bonds <- read.csv("credit_risk_data.csv")


# Collapsing categories in Sector.
bonds$Sector <- ifelse(bonds$Sector == "Brokerage/Asset Managers/Exchanges", "Finance", bonds$Sector)
bonds$Sector <- ifelse(bonds$Sector == "Electric", "Energy", bonds$Sector)
bonds$Sector <- ifelse(bonds$Sector == "Natural Gas", "Energy", bonds$Sector)
bonds$Sector <- ifelse(bonds$Sector == "Reits", "Finance", bonds$Sector)
bonds$Sector <- ifelse(bonds$Sector == "Transportation", "Communications", bonds$Sector)

# collapsing categories in BB_COMPOSITE
bonds$BB_COMPOSITE <- ifelse(bonds$BB_COMPOSITE %in% c("AAA", "AA+", "AA", "AA-"), ">A+", bonds$BB_COMPOSITE)
bonds$BB_COMPOSITE <- ifelse(bonds$BB_COMPOSITE %in% c('BB+', 'BBB-'), "<BBB", bonds$BB_COMPOSITE)
bonds$BB_COMPOSITE <- factor(bonds$BB_COMPOSITE, levels = c('>A+', 'A+', 'A', 'A-', 'BBB+', 'BBB', '<BBB'))

# Transform the 'bonds' column into an ordered factor
bonds$BB_COMPOSITE <- ordered(bonds$BB_COMPOSITE, levels = c(">A+", "A+", "A", "A-", "BBB+", "BBB",  "<BBB"))


bonds$Sector <- factor(bonds$Sector)
```

```{r, ignore = FALSE}
#| echo: false
library(MASS)
library(tidyverse)
library("car")
library(dplyr)
library(caret)
library(lmtest) # For likelihood ratio test
```

## Modeling

##### Technique

Given the ordinal nature of the outcome variable (Bloomberg Composite Credit Ratings), Ordinal Logistic Regression was chosen to answer the research question. Ordinal logistic regression is a statistical model used to analyze ordinal data, which is categorical data with an ordered ranking. Unlike nominal logistic regression, which is used for unordered categorical data, ordinal logistic regression takes into account the ordering of the categories.

##### Variable Selection

Once the A Priori variables were selected based on logical reasoning and domain knowledge, further statistical methods were used to drop some variable that do not add much value to the quality of the model.

## Full Model

```{r}
# full model
df <- bonds[bonds$Sector != "Banking", ]
ord_mod <- polr(BB_COMPOSITE ~ score + Sector + Duration + credit_spread + debt_to_assets + operating_profit_margin + ebitda_margin + roa + debt_to_equity, data=df, Hess=TRUE)
summary(ord_mod)
```

## Reduced Model

With the goal of making the final model easier to interpret and understand, less susceptible to over-fitting, and increased generality , Likelihood Ratio Tests were performed to prune statistically insignificant predictors.

```{r}
# Assuming original_df is your original data frame
df_r <- df[, c("BB_COMPOSITE", "score","Sector",  "Duration", "credit_spread", "debt_to_assets", "operating_profit_margin", "ebitda_margin", "roa")]

df_r <- na.omit(df_r)

# Define the reduced model (without the variable to be tested)
reduced_model <- polr(BB_COMPOSITE ~ score + Sector + Duration + credit_spread + debt_to_assets + operating_profit_margin + ebitda_margin + roa, data = df_r, Hess=TRUE)

# Perform likelihood ratio test
lr_test <- lrtest(ord_mod, reduced_model)

# Print the results
print(lr_test)
```

A Liklihood ratio test is performed above. The null hypothesis (H0) is that removing **`debt_to_equity`** does not significantly worsen the model fit. The chi-squared statistic is 0.1586 with 1 degree of freedom, resulting in a p-value of 0.6904. The p-value is greater than the typical significance level of 0.05, suggesting that there is not enough evidence to reject the null hypothesis. Therefore, based on the likelihood ratio test, removing the **`debt_to_equity`** variable does not significantly impact the model fit, and Model 2 may be preferred due to its simplicity.

```{r}
# Assuming original_df is your original data frame
df_r <- df[, c("BB_COMPOSITE", "score",  "Sector", "Duration", "credit_spread", "debt_to_assets", "operating_profit_margin", "ebitda_margin")]
df_r <- na.omit(df_r)

# Define the reduced model (without the variable to be tested)
reduced_model <- polr(BB_COMPOSITE ~ score + Duration + credit_spread + debt_to_assets + operating_profit_margin + ebitda_margin, data = df_r, Hess=TRUE)

# Perform likelihood ratio test
lr_test <- lrtest(ord_mod, reduced_model)

# Print the results
print(lr_test)
```

```{r}
table(bonds$BB_COMPOSITE)
table(bonds$Sector)
```

```{r}
pvals <- pnorm(-abs(summary(ord_mod)$coef[,"t value"]))*2
ctable <- cbind(summary(ord_mod)$coef,pvals)

ctable
```

```{r}
exp_coefs <- exp(cbind(OR=coef(ord_mod),confint(ord_mod)))
print(exp_coefs)
```

```{r}
vif(reduced_model)
```

```{r}
vif(ord_mod)
```

### Proportional Odds assumption

In ordinal regression models, the proportional odds assumption posits that the effect of predictors is consistent across different levels of the ordinal response variable. Testing the assumption of parallel lines is crucial in assessing the validity of the model.

```{r}
library(VGAM)
library(brant)

# Test for parallel lines assumption using brant
brant_test_result <- brant(reduced_model)

# Print the test result
print(brant_test_result)
```
